Cartesian Closed Categories

### 2017-09-21 papers/talk-8/CCC.tex
Fixed some paths

\def\pathToRoot{../}\input{\pathToRoot headers/uebungHeader}

@@

\end{definition}
\\
While it is required, that the reader knows about products, the idea of exponentials is assumed to be new and needs some more explanations.
\\
Regarding the products, just assess at this point, that the notion of all finite products is exhaustively described by terminal objects and binary products, because a terminal object represents the empty product, whereas every other product can be constructed with binary products. Since the whole concept of CCC is based on categories, whose functors are again objects, the existence of all finite products follows inductively, as long as it is clear that binary products are again objects.

@@

\section{Examples}
Now, we have what is necessary to look at a few examples to justify, that there is a need for the abstract concept of CCC. 
\subsection{\textbf{Sets}}
\begin{itemize}
\item \underline{\textit{Products:}} As already argued in the introduction, the cartesian product yields new objects $A\timesB\in ob(\textbf{Sets})$. Furthermore, it delivers the empty product $\emptyset \times \emptyset \to \emptyset$. Because all larger products can be build form binary products, \textbf{Sets} meets the first requirement to be a CCC. 
\item  \underline{\textit{Exponentials:}} As exponential, we choose the function space $\forall A, B \in ob(\textbf{Sets}), A^B := \{ f: A \to B| f functional\}$.
  \\ As proof obligation arises, that $\textbf{Sets}(A \times B, C ) \cong \textbf{Sets}(A, C^B), \forall A,B,C \in ob(\textbf{Sets})$ holds; that means that there are two arrows $f:A \times B \to C$ and $f':A \to C^B$ with $f (x,y) = f' (x) (y), \forall x \in A, y \in B$.
  \begin{itemize}
\item "$\Rightarrow$" : Assume $f: A \times B \to C$ with $f(x,y) \in A \times B \to C, x \in A, y \in B$.
\\ By fixing $x \in A$, one obtains an arrow $$f_x$$ with $$f_x(y) \in C^B.$$ By unfixing x, $f_x$ is morphed in the curried function $$\mu f:A\toC^B$$ with $$\mu f (x)(y) \in A \to C^B,$$ such that $$\mu f(x)(y)=f(x,y).$$ 
\item "$\Leftarrow$" : Assume $f': A \to C^B$. One simply defines a cartesian function $$g: A \times B \to C$$ as $$g(x,y) := f(x)(y), x \in A, y \in B$$ and is done.
  \end{itemize}
\end{itemize}

@@

  \subsection{\textbf{Posets}}
  The example of the category of preordered sets requires to exploit the evaluation arrow $\epsilon$ and the transposition $\lambda$ to verify the existence of exponential objects.
    \begin{itemize}
  \item \underline{\textit{Products:}}
    For all elements P, Q $\in ob(\textbf{Posets})$, the arrows are the monotone functions,
    \\ Hom(P,Q) := $\{f: P \to Q \mid f \quad  monotone \}$ with $\forall p,p', p\leq p' \to f p \leq f p'$.
  We can construct a binary product $P \times Q$, which is again a poset,using pairs $$(p,q) \in ob(P\timesQ)$$ and a partial ordering $$(p,q) \leq (p',q') \quad iff\quad p \leq p' \wedge q \leq q' .$$
  From this follow monotone projections: \\
\[
  \begin{tikzcd}
    & P && P\times Q \arrow{ll}{\pi_2} \arrow{rr}{\pi_1}  &&  Q\\
  \end{tikzcd}
\]
  And a monontone pairing
  $$ <f,g>: X \to P\times Q$$
  if $f:X \to P$ and $g: X\to Q$ are monotone.

\item \underline{\textit{Exponentials}}
  As exponential $Q^P$, we consider a set of monotone functions
  $$Q^P:=\{f:Q\toP \mid f \quad monotone \}$$
  with an pointwise ordering, such that $\forall f,g \in Q^P$,
  $$ f \leq g \leftrightarrow f p \leq g p,\forall p \in P.$$

  The evalution comes as an arrow
  $$\epsilon: Q^P\times P \to Q \quad with \quad \epsilon(f,p):=f(p)$$
  and the transposition is the arrow
  $$\lambda f : X \to Q^P \quad with \quad \lambda f(x) \in Q^P$$
  for some poset X and arrow
  $$f:X \times P \to Q \quad with \quad f(x,p):=\lambda f(x)(p)$$
  If these arrows proof to be monotone, $Q^P$ is an exponential object within \textbf{Posets}
  For this, we take some pairs $(f,p),(f',p')\in Q^P\times P, (f,p)\leq(f',p')$ and plug them into $\epsilon$:
  $$\epsilon(f,p) = f(p)$$
  $$\leq f(p')$$
  $$\leq f'(p') $$
  $$=\epsilon(f',p') $$
  Hence, $\epsilon$ is monotone. To show monotonicity of the transposition, we assume
  $f:X\timesP\to Q$ to be monotone and $x\leq x'$ for some $x,x' \in X$.
  $$\lambda f(x) \leq \lambda f(x') \in Q^P$$
  changes for any $p \in P$ to
  $$\lambda f(x)(p)\leq \lambda f(x')(p)$$
  which is by defintionß
  $$f(x,p)= \lambda f (x)(p) \leq \lambda f(x')(p) = f(x',p),$$
  so the proposition holds by equality $f(x,p)=\lambda f (x)(p)$ and assumption.
  \end{itemize}
  \newpage

@@

\\
The transpose $$a\leq b \Rightarrow c $$ is monotone as well, as shown by these inequations:
$$ a \land b \leq c \ to \neg b \lor (a \land b) \leq \neg b \lor c $$
$$ (\neg b \lor a) \land (\neg b \lor b) \leq b \Rightarrow c$$
$$ \neg b \lor a \leq (\neg b \lor a) \land (\neg b \lor b) $$
$$ a \leq \neg b \lor a$$
Hence
$$ a \leq b \Rightarrow c.
$$
\subsection{Counterexamples}

One striking property of any cartesian closed categories $\mathcal{A}$ with an initial object 0 and some object A is, that the isomorphism $0 \times A \cong 0$ holds,
because $$Hom(0 \times A, -) \cong Hom(0, (-)^A) \cong \{ \ast \}$$
where {$\ast$} denotes the one object, one arrow category, which is isomorphic to 0.
 \\
 So, to dismiss cartesian closedness for some category $\mathcal{B}$ with object A and an initial object C, it suffices to show, that $0 \times A \cong 0$ does not hold.
So in the case of $\boldsymbol{Vect_k}$ with the initial object 0:= ({x},$\lambda \cdot x=x$, x+x=x), the binary product is the direct product of two vectorspaces, but $0 \oplus A \cong A$. Therefore, $\boldsymbol{Vect_k}$ is not cartesian closed.
 \\
The identity above can be exploited to argue, that pointed categories, where the terminal object is identical to the initial object, are only CCC if they are trivial.
Because assuming \mathcal{A} to be a pointed category yields $$Hom(A,B) \cong Hom(1,A^B) \cong Hom(0,A^B) \cong 1$$  $\forall A,B \in ob(\mathcal{A})$.
With this argument, the category of relations \textbf{Rel} or of groups \textbf{Grp} can be ruled out to be a CCC, because both are pointed with the empty set, resp. the trivial group as zero objects.

@@

   \\ So we instantiate this with the category $\mathcal{A}$ having exponential objects $B^A, A,B \in ob(\mathcal{A})$ , $\epsilon: B^A \times A \to B$, and an invertible transpose $\lambda$.
   Lets define $$F(Z):=Z \times B, G(Z):= Z^B$$. 
   For g, we take $1_B$, for $\epsilon$ and have
   $$1_B \circ \epsilon: A \times B \to B \to B $$ with the transpose
   $$\lambda (1_B \circ \epsilon) : A \to (B \to B)^B = A \to B^B \to B^B$$ which is the type of
   $$(1_B)^B \circ \lambda\epsilon$$ what instantiates the first axiom as
   \[
   \begin{tikzcd}[column sep=normal]
     \lambda(A \times B \arrow{r}{\epsilon} & B \arrow{r}{1_B} & B) = A \arrow{r}{\lambda \epsilon} & B^B \arrow{r}{(1_B)^B} & B^B 
   \end{tikzcd}
   \]

>>>>

\def\pathToRoot{../../}\input{\pathToRoot headers/uebungHeader}

@@

\end{definition}

While it is required, that the reader knows about products, the idea of exponentials is assumed to be new and needs some more explanations.

Regarding the products, just assess at this point, that the notion of all finite products is exhaustively described by terminal objects and binary products, because a terminal object represents the empty product, whereas every other product can be constructed with binary products. Since the whole concept of CCC is based on categories, whose functors are again objects, the existence of all finite products follows inductively, as long as it is clear that binary products are again objects.

@@

\section{Examples}
Now, we have what is necessary to look at a few examples to justify, that there is a need for the abstract concept of CCC. 
\subsection{\textbf{Sets}}
\begin{itemize}
\item \underline{\textit{Products:}} As already argued in the introduction, the cartesian product yields new objects $A\times B\in ob(\textbf{Sets})$. Furthermore, it delivers the empty product $\emptyset \times \emptyset \to \emptyset$. Because all larger products can be build form binary products, \textbf{Sets} meets the first requirement to be a CCC. 
\item  \underline{\textit{Exponentials:}} As exponential, we choose the function space $\forall A, B \in ob(\textbf{Sets}), A^B := \{ f: A \to B| f functional\}$.
  \\ As proof obligation arises, that $\textbf{Sets}(A \times B, C ) \cong \textbf{Sets}(A, C^B), \forall A,B,C \in ob(\textbf{Sets})$ holds; that means that there are two arrows $f:A \times B \to C$ and $f':A \to C^B$ with $f (x,y) = f' (x) (y), \forall x \in A, y \in B$.
  \begin{itemize}
\item "$\Rightarrow$" : Assume $f: A \times B \to C$ with $f(x,y) \in A \times B \to C, x \in A, y \in B$.
\\ By fixing $x \in A$, one obtains an arrow $$f_x$$ with $$f_x(y) \in C^B.$$ By unfixing x, $f_x$ is morphed in the curried function $$\mu f:A\to C^B$$ with $$\mu f (x)(y) \in A \to C^B,$$ such that $$\mu f(x)(y)=f(x,y).$$ 
\item "$\Leftarrow$" : Assume $f': A \to C^B$. One simply defines a cartesian function $$g: A \times B \to C$$ as $$g(x,y) := f(x)(y), x \in A, y \in B$$ and is done.
  \end{itemize}
\end{itemize}

@@

  \subsection{\textbf{Posets}}
  The example of the category of preordered sets requires to exploit the evaluation arrow $\epsilon$ and the transposition $\lambda$ to verify the existence of exponential objects.
    \begin{itemize}
  \item \underline{\textit{Products:}}
    For all elements P, Q $\in ob(\textbf{Posets})$, the arrows are the monotone functions,
    \\ Hom(P,Q) := $\{f: P \to Q \mid f \quad  monotone \}$ with $\forall p,p', p\leq p' \to f p \leq f p'$.
  We can construct a binary product $P \times Q$, which is again a poset,using pairs $$(p,q) \in ob(P\times Q)$$ and a partial ordering $$(p,q) \leq (p',q') \quad iff\quad p \leq p' \wedge q \leq q' .$$
  From this follow monotone projections: \\
\[
  \begin{tikzcd}
    & P && P\times Q \arrow{ll}{\pi_2} \arrow{rr}{\pi_1}  &&  Q\\
  \end{tikzcd}
\]
  And a monontone pairing
  $$ <f,g>: X \to P\times Q$$
  if $f:X \to P$ and $g: X\to Q$ are monotone.

\item \underline{\textit{Exponentials}}
  As exponential $Q^P$, we consider a set of monotone functions
  $$Q^P:=\{f:Q\to P \mid f \quad monotone \}$$
  with an pointwise ordering, such that $\forall f,g \in Q^P$,
  $$ f \leq g \leftrightarrow f p \leq g p,\forall p \in P.$$

  The evalution comes as an arrow
  $$\epsilon: Q^P\times P \to Q \quad with \quad \epsilon(f,p):=f(p)$$
  and the transposition is the arrow
  $$\lambda f : X \to Q^P \quad with \quad \lambda f(x) \in Q^P$$
  for some poset X and arrow
  $$f:X \times P \to Q \quad with \quad f(x,p):=\lambda f(x)(p)$$
  If these arrows proof to be monotone, $Q^P$ is an exponential object within \textbf{Posets}
  For this, we take some pairs $(f,p),(f',p')\in Q^P\times P, (f,p)\leq(f',p')$ and plug them into $\epsilon$:
  $$\epsilon(f,p) = f(p)$$
  $$\leq f(p')$$
  $$\leq f'(p') $$
  $$=\epsilon(f',p') $$
  Hence, $\epsilon$ is monotone. To show monotonicity of the transposition, we assume
  $f:X\times P \to Q$ to be monotone and $x\leq x'$ for some $x,x' \in X$.
  $$\lambda f(x) \leq \lambda f(x') \in Q^P$$
  changes for any $p \in P$ to
  $$\lambda f(x)(p)\leq \lambda f(x')(p)$$
  which is by defintionß
  $$f(x,p)= \lambda f (x)(p) \leq \lambda f(x')(p) = f(x',p),$$
  so the proposition holds by equality $f(x,p)=\lambda f (x)(p)$ and assumption.
  \end{itemize}
  \newpage

@@

The transpose $$a\leq b \Rightarrow c $$ is monotone as well, as shown by these inequations:
$$ a \land b \leq c \ to \neg b \lor (a \land b) \leq \neg b \lor c $$
$$ (\neg b \lor a) \land (\neg b \lor b) \leq b \Rightarrow c$$
$$ \neg b \lor a \leq (\neg b \lor a) \land (\neg b \lor b) $$
$$ a \leq \neg b \lor a$$
Hence
$$ a \leq b \Rightarrow c.
$$
\subsection{Counterexamples}

One striking property of any cartesian closed categories $\mathcal{A}$ with an initial object 0 and some object A is, that the isomorphism $0 \times A \cong 0$ holds,
because $$Hom(0 \times A, -) \cong Hom(0, (-)^A) \cong \{ \ast \}$$
where {$\ast$} denotes the one object, one arrow category, which is isomorphic to 0.
 \\
 So, to dismiss cartesian closedness for some category $\mathcal{B}$ with object A and an initial object C, it suffices to show, that $0 \times A \cong 0$ does not hold.
So in the case of $\boldsymbol{Vect_k}$ with the initial object 0:= ({x},$\lambda \cdot x=x$, x+x=x), the binary product is the direct product of two vectorspaces, but $0 \oplus A \cong A$. Therefore, $\boldsymbol{Vect_k}$ is not cartesian closed.
 \\
The identity above can be exploited to argue, that pointed categories, where the terminal object is identical to the initial object, are only CCC if they are trivial.
Because assuming $\mathcal{A}$ to be a pointed category yields $$Hom(A,B) \cong Hom(1,A^B) \cong Hom(0,A^B) \cong 1$$  $\forall A,B \in ob(\mathcal{A})$.
With this argument, the category of relations \textbf{Rel} or of groups \textbf{Grp} can be ruled out to be a CCC, because both are pointed with the empty set, resp. the trivial group as zero objects.

@@

   So we instantiate this with the category $\mathcal{A}$ having exponential objects $B^A, A,B \in ob(\mathcal{A})$ , $\epsilon: B^A \times A \to B$, and an invertible transpose $\lambda$.
   Lets define $$F(Z):=Z \times B, G(Z):= Z^B$$. 
   For g, we take $1_B$, for $\epsilon$ and have
   $$1_B \circ \epsilon: A \times B \to B \to B $$ with the transpose
   $$\lambda (1_B \circ \epsilon) : A \to (B \to B)^B = A \to B^B \to B^B$$ which is the type of
   $$(1_B)^B \circ \lambda\epsilon$$ what instantiates the first axiom as
   \[
   \begin{tikzcd}[column sep=normal]
     \lambda(A \times B \arrow{r}{\epsilon} & B \arrow{r}{1_B} & B) = A \arrow{r}{\lambda \epsilon} & B^B \arrow{r}{(1_B)^B} & B^B 
   \end{tikzcd}
   \]
